---
title: "Data Visualization in R"
author: "Workshop Instructor"
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

In this exercise, we will visualize the cleaned and prepared data from **Exercise 1** using multiple R packages, including `ggplot2`, `echarts4r`, and `plotly`. This exercise assumes you have already completed the steps in **Exercise 1**, where we loaded, cleaned, and merged the data.

---

## Step 1: Load Necessary Libraries

To begin, let's load the required libraries for this exercise. These include visualization libraries (`ggplot2`, `echarts4r`, `plotly`) and other necessary packages.

```{r}
library(assertthat)
library(colorspace)
library(DT)
library(flexdashboard)
library(fontawesome)
library(glue)
library(knitr)
library(lubridate)
library(readr)
library(rio)
library(scales)
library(shiny)
library(shinyWidgets)
library(tidyverse)
library(echarts4r)
library(htmltools)
library(htmlwidgets)
library(plotly)
```

## Step 2: Define Color Scheme and Helper Function

We reuse the color scheme and helper function (`rnd_like_xl`) for consistency in the visualizations.

```{r}

load("gosa_data.RData")

# Define color scheme
color_scheme <- c('#66000040', '#022B3A40','#66666640')

# Excel-like rounding function
rnd_like_xl <- function(x, y) {
  round2 <- ifelse(x >= 0, round(x + 1e-09, y), round(x - 1e-09, y))
  return(round2)
}

# Define metro districts
metro_districts <- c(
  "Gwinnett County", "Cherokee County", "Cobb County", "DeKalb County", 
  "Douglas County", "Fayette County", "Clayton County", "Forsyth County", 
  "Fulton County", "Rockdale County", "Atlanta Public Schools", 
  "Buford City", "Decatur City", "Marietta City"
)
```

## Step 3: Visualizing 3rd Grade ELA %P/D by District

We will use **3rd Grade ELA** proficiency data to create a bar chart showing the top 50 districts with the highest proficiency and another focused on Metro Atlanta districts.

### Top 50 Districts in the State

```{r}     

df1 <- graph_data_sys |> filter(acdmc_lvl == '03')      

plt_title_1 <- unique(df1$plt_title)
plt_subtitle_1 <- unique(df1$plt_subtitle)

df1 |>
  arrange(desc(pct_prof_dis)) |> 
  slice_head(n = 50) |> 
  ggplot(
    aes(
      x = reorder(school_dstrct_nm, pct_prof_dis),
      y = pct_prof_dis,
      fill = sys_group,
      color = sys_group, 
      text = txt_subj
    )
  ) +
  geom_col() +
  scale_fill_manual(
    name = "School District Location:",
    values = color_scheme, 
    aesthetics = c('color', 'fill')
  ) +
  labs(
    title = plt_title_1, 
    subtitle = plt_subtitle_1, 
    x = " ", 
    y = "% Proficient or Distinguished"
  ) + 
  theme_minimal() + 
  theme(legend.position = "bottom") + 
  coord_flip()
```

### Metro Atlanta Districts

```{r}
df1 |>
  arrange(desc(pct_prof_dis)) |> 
  filter(school_dstrct_nm %in% c(metro_districts, "State of Georgia")) |> 
  ggplot(
    aes(
      x = reorder(school_dstrct_nm, pct_prof_dis),
      y = pct_prof_dis,
      fill = sys_group,
      color = sys_group, 
      text = txt_subj
    )
  ) +
  geom_col() +
  scale_fill_manual(
    name = "School District Location:",
    values = color_scheme, 
    aesthetics = c('color', 'fill')
  ) +
  labs(
    title = plt_title_1, 
    subtitle = plt_subtitle_1, 
    x = " ", 
    y = "% Proficient or Distinguished"
  ) + 
  theme_minimal() + 
  theme(legend.position = "bottom") + 
  coord_flip()
```

## Step 4: Scatter Plot of All Georgia Schools

We will create a scatter plot showing proficiency rates against the percentage of students directly certified.

```{r}
plt1 <- df1 |>
  ggplot(
    aes(
      x = direct_cert_perc,
      y = pct_prof_dis,
      fill = sys_group,
      color = sys_group, 
      text = txt_subj
    )
  ) +
  geom_point() +
  scale_fill_manual(
    name = "School District:",
    values = color_scheme, 
    aesthetics = c('color', 'fill')
  ) +
  labs(
    title = glue('{plt_title_1} by Direct Certification Rate'), 
    subtitle = plt_subtitle_1, 
    x = "% Direct Certification", 
    y = "% Proficient or Distinguished"
  ) + 
  theme(legend.position = "bottom")
```

### Static Plot

```{r}
# Static version of the scatter plot
plt1
```

### Interactive Plot using Plotly

```{r}
# Convert ggplot to interactive plotly
ggplotly(plt1, tooltip = c("x", "y", "text"))
```

## Step 5: Bar Chart using Echarts4r

We will use `echarts4r` to create an interactive bar chart comparing the proficiency rates of Metro Atlanta districts.

```{r}   

df1 |>
  arrange(pct_prof_dis) |>
  filter(school_dstrct_nm %in% c(metro_districts, "State of Georgia")) |> 
  e_charts(school_dstrct_nm) |>
  e_bar(pct_prof_dis) |>
  e_add_nested("itemStyle", color = color) |> 
  e_x_axis(
    name = "",
    nameLocation = "middle",
    nameTextStyle = list(color = "#333", fontSize = 14, fontStyle = 'bolder')
  ) |>
  e_y_axis(
    name = "\n% Proficient or Distinguished",
    max = 100,
    nameLocation = "middle",
    nameTextStyle = list(color = "#333", fontSize = 14, fontStyle = 'bolder')
  ) |>
  e_flip_coords() |> # flip axis
  e_title(
    text = "Milestones Proficiency Rate (%P/D): Grade 3 ELA EOG",
    subtext = "Metro Atlanta Public School Systems, School Year 2023-24",
    sublink = "https://www.gcpsk12.org/Page/24850",
    textStyle = list(color = "#333", fontWeight = "bolder")
  )
```

## Step 6: Scatter Plot using Echarts4r

We will create a scatter plot using `echarts4r`, focusing on the relationship between direct certification percentage and proficiency for school districts and schools.

### Scatter Plot for Districts

```{r}   

df1 |>
  group_by(factor(sys_group)) |>
  e_charts(direct_cert_perc) |>
  e_scatter(pct_prof_dis, bind = tooltip_lbl, symbol_size = 12) |>
  e_add_nested("itemStyle", color = color) |> 
  e_x_axis(
    name = "% Direct Certification",
    max = 100,
    interval = 10,
    nameLocation = "middle",
    nameTextStyle = list(fontSize = 14)
  ) |>
  e_y_axis(
    name = "% Proficient or Distinguished",
    max = 100,
    nameLocation = "middle",
    nameTextStyle = list(fontSize = 14)
  ) |>
  e_theme("infographic") |>
  e_legend(  
    selector = list(  
      list(type = 'all', title = 'Reset')
    )
  ) |> 
  e_color(c("#660000", "#022B3A", "#666666")) |> 
  e_title(
    text = "Milestones Proficiency Rate (%P/D): Grade 3 ELA EOG",
    subtext = "Georgia Public Schools, School Year 2023-24",
    sublink = "https://www.gcpsk12.org/Page/24850",
    textStyle = list(color = "#333", fontWeight = "bolder")
  ) |> 
  e_tooltip(
    axisPointer = list(
      type = "shadow"
    ), 
    valueDecimals = 1,
    formatter = htmlwidgets::JS("
    function(params) {
      var vals = params.name.split(',');
      var pointColor = params.color;
      var textColor = (pointColor === '#67b7d1' || pointColor === '#CCCCCC') ? '#FFFFFF' : '#000000';

      return (
        '<div style=\"font-family: Roboto, sans-serif; padding:

 5px; border: 2px solid ' + pointColor + ';\">' +
          '<div style=\"background-color:' + pointColor + '; padding: 5px;\">' +
            '<span style=\"color:' + textColor + '; font-size: 16px; font-weight: 700;\">' + vals[0] + '</span><br/>' +
            '<span style=\"color:' + textColor + '; font-size: 14px; font-weight: 400;\">' + vals[1] + '</span>' +
          '</div>' +
          '<br/><b>Actual Performance:</b> ' + params.value[1] +
          '<br/><b>Direct Certification:</b> ' + params.value[0] +
        '</div>'
      );
    }
    ")
  ) %>%
  e_toolbox_feature("dataZoom") %>%
  e_toolbox_feature(
    feature = "saveAsImage",
    title = "Save as image"
  )
```

### Scatter Plot for Schools

```{r}     


df2 <- graph_data |> filter(acdmc_lvl == '03')

df2 |>
  group_by(factor(sys_group)) |>
  e_charts(direct_cert_perc) |>
  e_scatter(pct_prof_dis, bind = tooltip_lbl, symbol_size = 12) |>
  e_add_nested("itemStyle", color = color) |> 
  e_x_axis(
    name = "% Direct Certification",
    max = 100,
    interval = 10,
    nameLocation = "middle",
    nameTextStyle = list(fontSize = 14)
  ) |>
  e_y_axis(
    name = "% Proficient or Distinguished",
    max = 100,
    nameLocation = "middle",
    nameTextStyle = list(fontSize = 14)
  ) |>
  e_theme("infographic") |>
  e_legend(  
    selector = list(  
      list(type = 'all', title = 'Reset')
    )
  ) |> 
  e_color(c("#660000", "#022B3A", "#666666")) |> 
  e_title(
    text = "Milestones Proficiency Rate (%P/D): Grade 3 ELA EOG",
    subtext = "Georgia Public Schools, School Year 2023-24",
    sublink = "https://www.gcpsk12.org/Page/24850",
    textStyle = list(color = "#333", fontWeight = "bolder")
  ) |> 
  e_tooltip(
    axisPointer = list(
      type = "shadow"
    ), 
    valueDecimals = 1
  ) %>%
  e_toolbox_feature("dataZoom") %>%
  e_toolbox_feature(
    feature = "saveAsImage",
    title = "Save as image"
  )    

```

---

## Conclusion

In this exercise, we visualized data from the previous exercise using both static and interactive plots. We employed libraries like `ggplot2`, `plotly`, and `echarts4r` to create insightful visualizations, such as bar charts and scatter plots, to display performance metrics for districts and schools in Georgia.
